Particle rejuvenation of Rao-Blackwellized Sequential Monte Carlo smoothers for Conditionally Linear and Gaussian models

### 2017-05-15 TexArxiv/answer_referees.tex

The paper surveys recent Rao-Blackwellized particle smoothers and presents a particle rejuvenation approach. The particle rejuvenation is interesting and seems useful, but it also shares some resemblance with the MCMC-based rejuvenation strategy proposed by [LBS+16, Section V.B] in the context of Rao-Blackwellized FFBS and with the strategy proposed by [FWT10] in the context of two-filter smoothing. Please relate the proposed method to these prior works. The detailed comments by the reviewers need to be addressed as well.

\vspace{.2cm}

{\em

It is true that our rejuvenation step shares some resemblance with the algorithms of [FWT10] and [LBS+16, Section V.B]. As noted in [FWT10, Secion~2.6], two-filter smoothers are prone to suffer from degeneracy issues when the algorithm associates forward particles at time $i-1$ with backward particles at time $i$. The authors illustrate this issue in the case where the hidden state is an AR(2) process. To overcome the weakness of such standard two-filter approaches, [FWT10] samples new particles at time $i$ conditional on forward particles up to time $i-1$ and on backward particles from time $i+1$ which are appropriately weighted. This allows to produce new particles at time $i$ and to obtain a SMC approximation whose support is not restricted to the backward particles at time $i+1$ as in the original two-filter algorithm.  The particle rejuvenation proposed in our paper exploits this idea in the specific case of linear and Gaussian models where explicit computations allows to produce an approximation using $(a^k_{1:i-1})_{1\le k \le N}$ and $(\tilde{a}^k_{i+1:n})_{1\le k \le N}$ with support $\{1,\ldots,J\}$. In addition, it produces a direct approximation of the target distribution based on $(a^k_{1:i-1})_{1\le k \le N}$ and $(\tilde{a}^k_{i+1:n})_{1\le k \le N}$  without any additional importance sampling steps.

On the other hand, another modification of the FFBS algorithm based on a Markov chain Monte Carlo sampling step was introduced in [LBS+16]. Instead of sampling from  explicitly from (12), [LBS+16] proposed to draw a forward path $a_{1:i-1}$ in $(a^k_{1:i-1})_{1\le k \le N}$ and a sate $a_i$ in $\{1,\ldots,J\}$ according to:
\[
\widetilde{q}(a_{1:i}|a_{i+1:n},y_{1:n}) = \sum_{k=1}^N \widetilde{\vartheta}^k_{i-1}\widetilde{q}(a_{i}|a^k_{1:i-1},a_{i+1:n},y_{1:n})\delta_{a^k_{1:i-1}}(a_{1:i-1})\eqsp,
\]
where $(\widetilde{\vartheta}^k_{i-1})_{1\le k \le N}$ are adjustment multipliers and $\widetilde{q}(a_{i}|a^k_{1:i-1},a_{i+1:n},y_{1:n})$ is a proposal kernel chosen by the user. This means that an ancestral path $a^{\star}_{1:i-1}$ is sampled in $(a^k_{1:i-1})_{1\le k \le N}$ with weights $(\widetilde{\vartheta}^k_{i-1})_{1\le k \le N}$ and $a^{\star}_i$ is sampled from $\widetilde{q}(\cdot|a^{\star}_{1:i-1},a_{i+1:n},y_{1:n})$. Then the proposed sequence $a^{\star}_{1:i}$ is accepted or rejected using the usual Metropolis-Hastings acceptance ratio. The choice of MCMC rejuvenation has interesting practical consequences as the computation of the acceptance ratio only requires to compute the posterior probability (11) for the proposed sequence $a^{\star}_{1:i}$ while our technique is based on the computation of  (11) for all combinations of sequences $(a^{k}_{1:i})_{1\le k \le N}$ and states $a_i\in\{1,\ldots,J\}$. The method proposed in [LBS+16] is computationally less costly as its complexity depends mainly on the number of MCMC samples produced by the algorithm. However it also requires a fair amount of tuning to choose efficient adjustment multipliers and proposal kernels. Our proposed method is computationally more intensive but allows to sample explicitly from (12) without additional tuning parameters. Note that both methods provide similar results for the same computational costs.
These remarks were added at the end of Section~2.2 and Section~2.3.1.
}

@@

\vspace{.5cm}

\noindent Some (more) intuition, about concepts and why certain steps are performed. Especially in the presentation of the RBTF in Section 2.4 I had problems to follow the overall idea. E.g. where do we integrate out the linear states in the backward pass of the original formulation presented by [BDM10]. I would expect this description before the modification of particle rejuvenation is presented. Also, some intuition about the explicit marginalization proposed in (9) would be helpful. Why is it chosen, what is the benefit compared to the original formulation? Currently, one gets easily lost in the derivations without maintaining the overall picture.

\vspace{.2cm}

{\em
\noindent  This comment helped to greatly improved the readability of the paper. As explained in the previous paragraph, Section~2.3.1 is now devoted to the original Rao-Blackwellized two-filter method and Section~2.3.2 to the rejuvenation step. The overall idea supporting two-filter based algorithms is clearly presented in Section~2.3.1 and most of the computations have also been postponed to the appendix in order to highlight the overall picture and avoid technicalities. Some additional motivations to introduce rejuvenation are given at the end of Sections~2.2.2 and~2.3.2 (see the answer to the associate editor). Two-filter smoothers are prone to suffer from degeneracy issues when the algorithm associates forward particles at time $i-1$ with backward particles at time $i$. The particle rejuvenation proposed in our paper exploits the idea introduced in [FWT10] in the specific case of linear and Gaussian models where explicit computations allows to produce an approximation using $(a^k_{1:i-1})_{1\le k \le N}$ and $(\tilde{a}^k_{i+1:n})_{1\le k \le N}$ with support $\{1,\ldots,J\}$ and without any additional sampling steps.
}

@@

\vspace{.5cm}

\noindent Some comparisons in terms of performance, complexity, detailing the superiority (or not) of one method over the other. E.g. what are the pros and cons of RB-FFBS and RBTF. Also some numerical comparisons in the simulation section are expected.

\vspace{.2cm}

{\em
\noindent Even is the paper is not a survey of all Rao-Blackwellized methods for CLGM anymore, some additional numerical experiments were added in Section~3. Each simulation study which was previously given in Section~3 was devoted either to FFBS or two-filter based methods. Now, both algorithms are used and compared to their counterpart with rejuvenation. Our main objective is then to show the improvement of the particle rejuvenation steps for both methods.  This also confirms that FFBS algorithm perform better even with a sharp choice for the  $\gamma_i$'s but also that rejuvenation steps improve both the accuracy and the variance of each method.
}

@@

\vspace{.5cm}

\noindent The minor comments were taken into consideration and typos were corrected. Some specific details are given below.

@@

\vspace{.5cm}

\noindent The authors talk about two FFBS algorithms that they compare to the FFBS with
particle rejuvenation. But which two? Kim's and Lindsten's? Some more details
would be helpful, also which forward filter has been used.

\vspace{.2cm}

{\em 
\noindent The two FFBS algorithms are the original version of [LBS+16]  and the algorithm with rejuvenation.
A first (costly) run of the FFBS algorithm with rejuvenation with a great number of particles ($N=5000$) is used as a benchmark value. Then the posterior distributions are estimated by the original methods and the methods with rejuvenation with fewer particles and compared to this  benchmark run.
}

\noindent Figure 1 shall be Table 1? In the text it is 2500 iterations, in the Table
caption it is 3500? How is Std. Dev obtained? I mean is this simulated data or real data where
the true parameters $\theta$ are usually unknown?

\noindent {\em 
The estimated values were indeed obtained after 2500 iterations of the algorithm. The standard deviations are obtained with 50 independent Monte Carlo runs of the estimation algorithm (and the estimated value is set as the mean over all runs). This is now stated clearly in the paper.
}

@@

\vspace{.2cm}

\noindent {\em 
The estimated values were indeed obtained after 2500 iterations of the algorithm. The standard deviations are obtained with 50 independent Monte Carlo runs of the estimation algorithm (and the estimated value is set as the mean over all runs). This is now stated clearly in the paper.
}

\subsubsection*{Referee 2}
\noindent It is well known that SMC algorithms, in their most basic form, are numerically unstable
when used for approximating distributions defined on path spaces of increasing dimension. This is due to the particle path degeneracy phenomenon caused by the resampling mechanism of SMC algorithms. The degeneracy appears to be a potential problem for the estimators discussed in the paper, since the approximation of the
backward kernel (on, e.g., p. 8, line 52) involves quantities such as the mean and the
covariance matrix of the conditional distribution of $z_i$ given the states $a_{1:i}$ and the observations,
evaluated along the particle genealogical paths. I miss a discussion on this. If the effect exists, does the rejuvenation proposed by the authors counteract the same? (I do not think so).

\vspace{.2cm}
{\em
\noindent As explained to the associate editor and to  Referee~1, it is true that particle smoothers are prone to degenerate when they are based on  the genealogy of the forward or backward particles. In addition, in the specific case of two-filter based smoothers, [FWT10, Section~2.6] noted that these algorithms are prone to suffer from degeneracy issues when the algorithm associates forward particles at time $i-1$ with backward particles at time $i$. 
The particle rejuvenation proposed in our paper produces an approximation using $(a^k_{1:i-1})_{1\le k \le N}$ and $(\tilde{a}^k_{i+1:n})_{1\le k \le N}$ with support $\{1,\ldots,J\}$. In the specific case of linear and Gaussian models, Kalman filtering techniques allow to compute explicitly the approximation of the target distribution at time $i$ based on  $(a^k_{1:i-1})_{1\le k \le N}$ and $(\tilde{a}^k_{i+1:n})_{1\le k \le N}$.

Therefore, the support of the approximated distribution is not restricted to $(\tilde{a}^k_{i})_{1\le k \le N}$ to overcome particle depletion. The marginalization over the linear states  builds new trajectories of the form $(a^k_{1:i-1},j,\tilde{a}^\ell_{i+1:n})_{1\le k,\ell \le N}$ for all $1\le j \le J$ which have not been sampled by the forward and backward filters. Although this does not overcome the overall degeneracy issue, numerical experiments highlight that this improves the performance of all Rao-Blackwellized SMC smoothers. Additional comments were added at the end of Section~2.2 and Section~2.3.1.
}

@@

\vspace{.5cm}

\noindent The rejuvenation approach proposed in Section 2.3 makes sense in that it allows the
production of backward draws that are not restricted to the discrete support formed by the particles generated in the forward  filtering pass. On the other hand, it is not obvious that there is a lot of variance to be gained in doing this (even though the simulation study in Section 3 indicates some improvement). Thus, I miss a heuristic
motivation for the rejuvenation step in Section 2.3. Are there specific models for which the rejuvenation step is supposed improve significantly the FFBS algorithm? As far as I understand, there is a price to pay for the rejuvenation in terms of some additional computational work, as the Kalman filter needs to be run for all the possible extensions
of each particle path. So, is the particle rejuvenation worthy of this additional effort? I guess it depends, e.g., on the number J of possible regimes?

\vspace{.2cm}
{\em
\noindent The intuition supporting particle rejuvenation is explained in the answers to the associate editor and to Referee~1. There is no theoretical result to prove that this approach provides a better approximation of the target distributions. However, our numerical experiments hint that this actually improves the accuracy and variance of the proposed estimators which would be even more interesting with additional regimes or additional rejuvenation steps. 
[FWT10, Section~2.6] illustrated the degeneracy issue of two-filter based smoothers with the example of an AR(2) process. Our rejuvenation exploits the idea introduced in [FWT10, Section~2.6] to overcome degeneracy issues in the context of CLGM. In addition, the additional computational cost is not prohibitive as Kalman filtering steps are much faster than the combination of the forward and backward particles whose complexity grows with $N^2$ (but who is stil necessary when no rejuvenation step is introduced).
}

@@

\vspace{.5cm}

\noindent It took a while before I understood the backward filter algorithm in Section 2.4. More
specifically, the expression of the importance weights on p. 11 seems to be based on
the identity
\[
\tilde{p}_{i}(a_{i:n}|y_{i:n}) \propto \left\{\prod_{u=i}^{n-1}Q(a_u,a_{u+1})\right\} \int \gamma^{}_{i}(a_i,z_i) p(y_{i:n}|z_i,a_{i:n})\rmd z_i\eqsp.
\]
which never appears explicitly in the description of the backward  filter. I would recommend
the authors to include a derivation of the backward filter on the basis of this identity.

\vspace{.2cm}
{\em
\noindent The identity was added in Section~2.3.1 in a more detailed derivation of the backward filter. We thank Referee~2 for this help to improve the presentation of the backward filter.
}

@@

\vspace{.5cm}

\noindent Finally, I miss a discussion on the last sampling operation of the Rao-Blackwellized
two-filter smoother, i.e., the sampling of the mixture obtained when plugging the approximations into (9). This should yield a mixture with $N^2$ components, implying a not innocent computational complexity when N is large.

\vspace{.2cm}
{\em
\noindent The additional computational cost of the rejuvenation step comes from the Kalman filtering steps needed to extend all backward trajectories (built with particles). These steps are much faster than the combination of the forward and backward particles whose complexity grows with $N^2$. However, this combination is also necessary in the original two-filter algorithm of [BDM10] with no rejuvenation. A comment was added in the numerical section.
}

### 2017-05-15 TexArxiv/clgm.tex
Final modifications for revision

%\paragraph{Experiment 2:}
%In this second experiment, we use the same parameters except that we increase by a factor $10$ the values in $\barG$. The smoothing distributions are estimated with $N=15$ particles. The specific form of $B_1$ (first row with entries equal to one) and the fact that $T_1 = T_2$, $B_1 = B_2 $ and $\barG_1 = \barG_2$ are also motivated by the application to WTI crude oil data.  It is assumed that the differences between the two regimes are given by the state's dynamics ($d_1$, $d_2$, $\barH_1$ and $\barH_2$). Figures~\ref{fig:exp1:label} and~\ref{fig:exp2:label} displays the approximations of the  posterior distributions of the regimes given by the algorithms using $\gamma^{}_i (a_i, z_i) = p^N(a_{i},z_{i}|y_{1:i-1})$ where $p^N(a_{i},z_{i}|y_{1:i-1})$ is defined by \eqref{eq:rN}.
%%p^N(a_{i},z_{i}|y_{1:i-1}) = \sum_{k=1}^N \omega_{\mathsf{f},i}^k(a_{i})\exp\left\{-\frac{1}{2}\normMat{\Omega^k_{i-1}(a_{i})}{z_{i}}^2+z'_{i}r_{i-1}^k(a_{i})\right\}\eqsp.
%%\end{equation}:
%%\[
%%\gamma^{}_i (a_i, z_i) = \sum_{k=1}^{N} Q(a^{k}_{i-1},a_i) \omega^{k}_{i-1} \exp\left\{ -\frac{1}{2} \log \left|\barH_{a^k_{i-1}} \right| -\frac{1}{2}\normMat{\barH_{a^k_{i-1}}}{z_i - d_{a^k_{i-1}} -T \mu^k_{i-1} }^2\right\} \eqsp.
%%\]

@@

\subsection{Model}
Modeling commodity prices is a crucial step to valuate contingent claims related to energy markets and to optimize storage or extraction strategies. In \cite{gibson:schwartz:1990,schwartz:1997}, the authors proposed a model where the spot price of a commodity$(S_t, t \geq 0)$ depends on a second factor $(\delta_t, t \geq 0)$, referred to as the instantaneous convenience yield. This factor plays the role of dividends in equity markets and models the benefit of holding the physical commodity or the storage and maintenance costs required to keep the commodity. In this model, this convenience yield is described as an Ornstein-Uhlenbeck process:
\begin{align*}
\rmd S_{t} & = (r-\delta_{t})S_{t}\rmd t+\sigma S_{t} \rmd W_{t}^{1} \eqsp,\\
\rmd \delta_{t} & = \kappa(\alpha-\delta_{t})\rmd t+\eta \rmd W_{t}^{2} \eqsp, \quad \rmd \langle W_t^1, W_t^2\rangle = \rho \rmd t\eqsp,
\end{align*}
where the parameter $\left(r, \sigma, \kappa, \alpha, \eta, \rho \right)$ are constant and $( (W_t^1,W_t^2), t \geq 0)$ are standard Brownian motions. This model appear to be too restrictive as energy markets are not likely to revert to a single equilibrium value. This assumption is relaxed using Markov switching models to allow several possible regimes for the spot price and the convenience yield. Following \cite{almansour:2016}, the spot price and convenience yield are described in this paper as:
\begin{align*}
\rmd S_{t} & = (r-\delta_{t})S_{t}\rmd t+\sigma_{a_t} S_{t}\rmd W_{t}^{1} \eqsp,\\
\rmd\delta_{t} & = \kappa (\alpha_{a_t} -\delta_{t})\rmd t+\eta_{a_t} \rmd W_{t}^{2} \eqsp, \quad \rmd\langle W_t^1, W_t^2\rangle = \rho_{a_t} \rmd t\eqsp,
\end{align*}
where $(a_t)_{t\ge 0}$ is a finite state space Markov process.

This model allows to exhibit fundamental features of commodity future prices, which typically display different regimes of volatility and/or convenience yield. A two-regime model is already sufficient to produce stylized effects such as contango (increase of future prices) and backwardation (decrease of future prices).  Assuming that the switching rate between regimes is negligible compared to the inverse of the discretization period, the discretized version of the spot price and convenience yield $Z_i= (\ln S_i, \delta_i)$ (the sampling period is taken to be 1) is modeled as a CLGM. The explicit integration of this SDE detailed in Lemma~\ref{lem:integratedSDE} yields the following discrete time model for $(Z_i)_{i\ge 0}$:

@@

All the conditional expectations involved in $Q(\theta,\theta_p)$ are approximated using our RBTF algorithm to define the SMC approximation  $\theta\mapsto Q^N(\theta,\theta_p)$ of $\theta\mapsto Q(\theta,\theta_p)$. As the function $\theta\mapsto Q^N(\theta,\theta_p)$ cannot be maximized analytically, the M-step is performed numerically using the Covariance Matrix Adaptation Evolution Strategy (CMA-ES) introduced in \cite{hansen:ostermeier:2001}. This derivative-free optimization procedure is known to perform well in complex multimodal optimization settings, see e.g. \cite{hansen:kern:2004}.

@@

In Gibson-Schwartz model \cite{gibson:schwartz:1990}, a stronger backwardation effect implies a greater value for $\alpha$ for the same values of the other parameters. For the CME WTI Crude Oil, backwardation effect is more frequent than contango effect so that $\alpha_1$ should be greater than $\alpha_2$. Therefore, this condition is imposed for all simulations in the CMA-ES algorithm. The results after $2500$ iterations of the EM algorithm are given in Table~\ref{fig:CMEfinalestimates}.
\begin{table}[h!]
	\centering
	\begin{tabular}{||c|c|c|c|c|c|c|c|c||c|clc|c|c||}
	\hline
		Parameter&  $\kappa$ & $\sigma_1$ & $\sigma_2$ & $\eta_1$ & $\eta_2$ & $\rho_1$ & $\rho_2$  \\
		\hline
		Value & 2.6378 &0.3733 & 0.3485 & 0.5892 & 0.3814 & 0.8709 & 0.6761   \\
		\hline
		Std. Dev & 0.1999 & 0.005438 & 0.002884 & 0.0483 & 0.0349 & 0.0064 & 0.0052 \\  
		\hline
	\end{tabular}

@@

	\begin{tabular}{||c|c|c|c|c|c|c|c|c|c|c|||}
	\hline
		Parameter & $\alpha_1$ & $\alpha_2$  & $g_1$ & $g_2$ & $g_3$ & $g_4$  & $Q(1,1)$ & $Q(2,2)$\\
		\hline
		Value & 0.0889 & -0.0281  & 2.3e-2 & 1.0e-4 & 3.0e-4 & 2.3e-2 & 0.9917 & 0.9880 \\
		\hline
		Std. Dev & 0.004248 & 0.00149 & 1.9e-4 & 2.6e-4 & 2.3e-4 & 2.1e-04 &  6.7e-4 & 9.6e-4\\
		\hline
	\end{tabular}
	\caption{Final estimates after 2500 iterations.}
	\label{fig:CMEfinalestimates}
\end{table}
As expected, we obtain $\sigma_1 \geq \sigma_2$, $\alpha_1 \geq \alpha_2$, $\eta_1 \geq \eta_2$ and $\rho_1 \geq \rho_2$ at convergence of the EM algorithm. Moreover, $Q(1,1) > Q(2,2)$ is corresponding to the prediction that we did from the data description. From the results, $\sigma_1 \geq \sigma_2$, $\alpha_1 \geq \alpha_2$ indicates the first regime (backwardation) characterized by a higher value in both volatility and equilibrium level of convenience yield, and the second regime (contango) characterized by a lower value in both volatility and equilibrium convenience yield level. This in accordance with the theory of storage that the volatility of the commodity spot price is high when the inventory is low, and the convenience yield is all the higher as inventory is low.

Figure~\ref{fig:LogPriceTermStructure:data} compares the evolution of future $1M$ (the nearest contracts) to the \emph{term structure} observed from CME WTI crude oil, defined as the difference of future $13M$ and future $1M$ (to avoid seasonality). The figure shows that it is not necessary to have an inverse relationship between the  price of the nearest contract and the term structure. But when a significant drop in the price  of the nearest contract occurs, the term structure increases (i.e. in contango).
\begin{figure}
\centering
\includegraphics[scale=.6,angle=90]{LogPrice_TermStructure}
\caption{Log-price (red line) and slope of future curves (blue line)}
\label{fig:LogPriceTermStructure:data}
\end{figure}

The correlation between the spot price and the convenience yield is positive and high in both two regimes. This is an accordance to what as been observed in most commodity market; see \cite{gibson:schwartz:1990}.  The slope of future curve decreases in function of maturity.

@@

\begin{figure}
	\centering
	\includegraphics[scale=.75,angle=90]{Posterior_Data}
	\caption{Posterior probability (triangle black) and slope of future curves (blue line)}
	\label{fig:posterior:data}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[scale=.75]{Posterior_Data_LogPrice}
	\caption{Posterior probability (triangle black) and slope of future curves (blue line)}
	\label{fig:posterior:data:logprice}
\end{figure}

\section{Conclusions}
This paper presents Rao-Blackwellized Sequential Monte Carlo methods to approximate smoothing distributions in conditionally linear and Gaussian state spaces in a common unifying framework and also provides a survey of different techniques that could be used in the forward filtering pass to improve significantly the usual mixture Kalman filter. The filtering distributions are approximated at  each time step by considering all possible offsprings of all ancestral trajectories before discarding degenerated paths instead of resampling the ancestral paths before propagating them at the next time step. We also investigated the benefit of additional Rao-Blackwellization steps to sample new regimes at each time step conditional on the forward and backward particles. This rejuvenation step uses explicit integration of the hidden linear states before merging the forward and backward filters for two-filter based algorithms or before sampling new states backward in time for FFBS based methods.  The paper displays some Monte Carlo experiments with simulated data to illustrate that this additional rejuvenation step improves the performance of the smoothing algorithms with no additional computational costs or variance. They are also applied to commodity markets using WTI crude oil data.

>>>>

%%
%%\paragraph{Experiment 2:}
%%In this second experiment, we use the same parameters except that we increase by a factor $10$ the values in $\barG$. The smoothing distributions are estimated with $N=15$ particles. The specific form of $B_1$ (first row with entries equal to one) and the fact that $T_1 = T_2$, $B_1 = B_2 $ and $\barG_1 = \barG_2$ are also motivated by the application to WTI crude oil data.  It is assumed that the differences between the two regimes are given by the state's dynamics ($d_1$, $d_2$, $\barH_1$ and $\barH_2$). Figures~\ref{fig:exp1:label} and~\ref{fig:exp2:label} displays the approximations of the  posterior distributions of the regimes given by the algorithms using $\gamma^{}_i (a_i, z_i) = p^N(a_{i},z_{i}|y_{1:i-1})$ where $p^N(a_{i},z_{i}|y_{1:i-1})$ is defined by \eqref{eq:rN}.
%%%p^N(a_{i},z_{i}|y_{1:i-1}) = \sum_{k=1}^N \omega_{\mathsf{f},i}^k(a_{i})\exp\left\{-\frac{1}{2}\normMat{\Omega^k_{i-1}(a_{i})}{z_{i}}^2+z'_{i}r_{i-1}^k(a_{i})\right\}\eqsp.
%%%\end{equation}:
%%%\[
%%%\gamma^{}_i (a_i, z_i) = \sum_{k=1}^{N} Q(a^{k}_{i-1},a_i) \omega^{k}_{i-1} \exp\left\{ -\frac{1}{2} \log \left|\barH_{a^k_{i-1}} \right| -\frac{1}{2}\normMat{\barH_{a^k_{i-1}}}{z_i - d_{a^k_{i-1}} -T \mu^k_{i-1} }^2\right\} \eqsp.
%%%\]

@@

\subsection{Model}
Modeling commodity prices is a crucial step to valuate contingent claims related to energy markets and to optimize storage or extraction strategies. In \cite{gibson:schwartz:1990,schwartz:1997}, the authors proposed a model where the spot price of a commodity$(S_t, t \geq 0)$ depends on a second factor $(\delta_t, t \geq 0)$, referred to as the instantaneous convenience yield. This factor plays the role of dividends in equity markets and models the benefit of holding the physical commodity or the storage and maintenance costs required to keep the commodity. In this model, this convenience yield is described as an Ornstein-Uhlenbeck process:
\rmd S_{t} & = (r-\delta_{t})S_{t}\rmd t+\sigma S_{t} \rmd W_{t}^{1} \eqsp,\\
\rmd \delta_{t} & = \kappa(\alpha-\delta_{t})\rmd t+\eta \rmd W_{t}^{2} \eqsp, \quad \rmd \langle W_t^1, W_t^2\rangle = \rho \rmd t\eqsp,
where the parameter $\left(r, \sigma, \kappa, \alpha, \eta, \rho \right)$ are constant and $( (W_t^1,W_t^2), t \geq 0)$ are standard Brownian motions. This model appears to be too restrictive as energy markets are not likely to revert to a single equilibrium value. This assumption is relaxed using Markov switching models to allow several possible regimes for the spot price and the convenience yield. Following \cite{almansour:2016}, the spot price and convenience yield are described in this paper as:
\rmd S_{t} & = (r-\delta_{t})S_{t}\rmd t+\sigma_{a_t} S_{t}\rmd W_{t}^{1} \eqsp,\\
\rmd\delta_{t} & = \kappa (\alpha_{a_t} -\delta_{t})\rmd t+\eta_{a_t} \rmd W_{t}^{2} \eqsp, \quad \rmd\langle W_t^1, W_t^2\rangle = \rho_{a_t} \rmd t\eqsp,
where $(a_t)_{t\ge 0}$ is a finite state space Markov process. This model allows to exhibit fundamental features of commodity future prices, which typically display different regimes of volatility and/or convenience yield. A two-regime model is already sufficient to produce stylized effects such as contango (increase of future prices) and backwardation (decrease of future prices).  Assuming that the switching rate between regimes is negligible compared to the inverse of the discretization period, the discretized version of the spot price and convenience yield $Z_i= (\ln S_i, \delta_i)$ (the sampling period is taken to be 1) is modeled as a CLGM. The explicit integration of this SDE detailed in Lemma~\ref{lem:integratedSDE} yields the following discrete time model for $(Z_i)_{i\ge 2}$:

@@

All the conditional expectations involved in $Q(\theta,\theta_p)$ are approximated using our two-filter algorithm with rejuvenation algorithm to define the SMC approximation  $\theta\mapsto Q^N(\theta,\theta_p)$ of $\theta\mapsto Q(\theta,\theta_p)$. As the function $\theta\mapsto Q^N(\theta,\theta_p)$ cannot be maximized analytically, the M-step is performed numerically using the Covariance Matrix Adaptation Evolution Strategy (CMA-ES) introduced in \cite{hansen:ostermeier:2001}. This derivative-free optimization procedure is known to perform well in complex multimodal optimization settings, see e.g. \cite{hansen:kern:2004}.

@@

In Gibson-Schwartz model \cite{gibson:schwartz:1990}, a stronger backwardation effect implies a greater value for $\alpha$ for the same values of the other parameters. For the CME WTI Crude Oil, backwardation effect is more frequent than contango effect so that $\alpha_1$ should be greater than $\alpha_2$. Therefore, this condition is imposed for all simulations in the CMA-ES algorithm. The results after $2500$ iterations of the EM algorithm are given in Table~\ref{fig:CMEfinalestimates}. The estimated values and standard deviations are obtained with $50$ independent runs of the algorithm.
\begin{table}[h!]
	\centering
	\begin{tabular}{||c|c|c|c|c|c|c|c|c||c|clc|c|c||}
	\hline
		Parameter&  $\kappa$ & $\sigma_1$ & $\sigma_2$ & $\eta_1$ & $\eta_2$ & $\rho_1$ & $\rho_2$  \\
		\hline
		Value & 2.6378 &0.3733 & 0.3485 & 0.5892 & 0.3814 & 0.8709 & 0.6761   \\
		\hline
		Std. Dev & 0.1999 & 0.005438 & 0.002884 & 0.0483 & 0.0349 & 0.0064 & 0.0052 \\  
		\hline
	\end{tabular}

@@

	\begin{tabular}{||c|c|c|c|c|c|c|c|c|c|c|||}
	\hline
		Parameter & $\alpha_1$ & $\alpha_2$  & $g_1$ & $g_2$ & $g_3$ & $g_4$  & $Q(1,1)$ & $Q(2,2)$\\
		\hline
		Value & 0.0889 & -0.0281  & 2.3e-2 & 1.0e-4 & 3.0e-4 & 2.3e-2 & 0.9917 & 0.9880 \\
		\hline
		Std. Dev & 0.004248 & 0.00149 & 1.9e-4 & 2.6e-4 & 2.3e-4 & 2.1e-04 &  6.7e-4 & 9.6e-4\\
		\hline
	\end{tabular}
	\caption{Final estimates after 2500 iterations.}
	\label{fig:CMEfinalestimates}
\end{table}
As expected, we obtain $\sigma_1 \geq \sigma_2$, $\alpha_1 \geq \alpha_2$, $\eta_1 \geq \eta_2$ and $\rho_1 \geq \rho_2$ at convergence of the EM algorithm. Moreover, $Q(1,1) > Q(2,2)$ corresponds to the prediction that we did from the data description. The fact that $\sigma_1 \geq \sigma_2$ and $\alpha_1 \geq \alpha_2$ indicates the first regime (backwardation) characterized by a higher value in both volatility and equilibrium level of convenience yield, and the second regime (contango) characterized by a lower value in both volatility and equilibrium convenience yield level. This in accordance with the theory of storage that the volatility of the commodity spot price is high when the inventory is low, and the convenience yield is all the higher as inventory is low.

Figure~\ref{fig:LogPriceTermStructure:data} compares the evolution of future $1M$ (the nearest contracts) to the \emph{term structure} observed from CME WTI crude oil, defined as the difference of future $13M$ and future $1M$ (to avoid seasonality). The figure shows that it is not necessary to have an inverse relationship between the  price of the nearest contract and the term structure. But when a significant drop in the price  of the nearest contract occurs, the term structure increases (i.e. in contango).
\begin{figure}
\centering
\includegraphics[scale=.6,angle=90]{LogPrice_TermStructure}
\caption{Log-price (red line) and slope of future curves (blue line).}
\label{fig:LogPriceTermStructure:data}
\end{figure}

The correlation between the spot price and the convenience yield is positive and high in both two regimes. This is an accordance to what as been observed in most commodity market, see \cite{gibson:schwartz:1990}.  The slope of future curve decreases in function of maturity.

@@

\begin{figure}
	\centering
	\includegraphics[scale=.75,angle=90]{Posterior_Data}
	\caption{Posterior probability (triangle black) and slope of future curves (blue line).}
	\label{fig:posterior:data}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[scale=.75]{Posterior_Data_LogPrice}
	\caption{Posterior probability (triangle black) and slope of future curves (blue line).}
	\label{fig:posterior:data:logprice}
\end{figure}

\section{Conclusions}
This paper presents Rao-Blackwellized Sequential Monte Carlo methods to approximate smoothing distributions in conditionally linear and Gaussian state spaces in a common unifying framework. It also provides different techniques that could be used in the forward filtering pass to improve significantly the usual mixture Kalman filter. The filtering distributions are approximated at  each time step by considering all possible offsprings of all ancestral trajectories before discarding degenerated paths instead of resampling the ancestral paths before propagating them at the next time step. The paper investigates the benefit of additional Rao-Blackwellization steps to sample new regimes at each time step conditional on the forward and backward particles. This rejuvenation step uses explicit integration of the hidden linear states before merging the forward and backward filters for two-filter based algorithms or before sampling new states backward in time for FFBS based methods.  The paper displays some Monte Carlo experiments with simulated data to illustrate that this additional rejuvenation step improves the performance of the smoothing algorithms with no substantial additional computational costs. They are also applied to commodity markets using WTI crude oil data.
